---
format: 
  html:
    page-layout: full
---

# Adopting a JAGS sampler
NIMBLE [de Valpine et al. 2017](https://www.tandfonline.com/doi/full/10.1080/10618600.2016.1172487) is a relatively new framework for fitting hierarchical models with MCMC. Previously, latent variable models were often fitted with JAGS [Plummer 2004](https://pdfs.semanticscholar.org/837b/9203abc8b3416e620d4c99d8500b4bd9be20.pdf) instead. An example for latent variable models, is the R-package Boral [Hui 2016](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1111/2041-210X.12514). 

Boral allows users to straightforwardly, in familiar R interface, specify a latent variable model. This model specification is then translated to BUGS code (which JAGS uses) using some text manipulation. A while ago I decided that NIMBLE can probably sample the parameters for such a model more efficiently, as it allows for more flexibility in sampler specification and blocking schemes for parameters. Off I went; translating the script outputted by Boral in a format that can be read by NIMBLE was straightforward. Unfortunately, relative to NIMBLE's automated factor slice sampler, JAGS provided much more effective samples than NIMBLE.

One of the reasons for this, I suspect, if that JAGS has a larger library of MCMC samplers. Especially when it comes to models that can be represented as a GLM. This is also the case for latent variable models, because those are (almost) GLMs when keeping the latent variable fixed, and while estimating the other parameters. As a fairer comparison, I have set off to code up some GLM samplers from JAGS in NIMBLE. This is quite a challenge, since JAGS's samplers are coded in C++, and, well, my C++ is rusty at best. Fortunately, the [pogit R-package](https://cran.r-project.org/web/packages/pogit/index.html) includes chunks of the samplers in R-code, so I can use that when my understanding of C++ fails (and to verify that everything works correctly). It also does slab-and-spike variable selection, but I can ignore that bit for now.

## The Sampler

JAGS has not one GLM sampler. It might seem like that at first glance because JAGS's "GLM factory" has multiple scripts (such as [this one](https://github.com/todesking/JAGS-code/blob/master/src/modules/glm/samplers/GLMMethod.cc)) specific to GLMs. However, the idea is that when fed a model (script), JAGS first checks if that model can be represented as a linear model. If it can, it checks the distribution for the responses and is fed to a corresponding sampler for that distribution. Consequently, the "GLM sampler" is in fact a multitude of samplers. Each sampler is based on the same idea; the GLM can be represented as a linear model with some smart statistical voodoo.

For the Poisson case this is based on the work by [Fruhwirth-Schnatter and Wagner 2004](https://research.wu.ac.at/en/publications/data-augmentation-and-gibbs-sampling-for-regression-models-of-sma-7) and [Fruhwirth-Schnatter et al. 2009](https://link.springer.com/article/10.1007/s11222-008-9109-4).

The sampler relies on representing Poisson random variables in terms of their inter-arrival times, and approximating the distribution of the inter-arrival times by mixtures of normal distributions. Conditional on some quantities, the Poisson regression can be (approximately) represented as a linear regression. On each iteration of the sampler, the parameters are (conditionally) sampled from a multivariate normal distribution. Consequently, we can think of the sampler (algorithm) in three steps: 1) update parameters conditional on the inter-arrival times $\boldsymbol{\tau}_i$ and indicator variables $\boldsymbol{R}_1$ and $\boldsymbol{R}_2$. These latter quantities are treated as missing data.

The *pogit* R-package has functions to update these quantities; [**iams1_poisson**](https://github.com/cran/pogit/blob/b0f4b72eaabb1ce32efa1f4ce5517be7ece2afd0/R/dataug_pois_iams.R#L69) updates $\boldsymbol{\tau}$, [**iams2_poisson**](https://github.com/cran/pogit/blob/b0f4b72eaabb1ce32efa1f4ce5517be7ece2afd0/R/dataug_pois_iams.R#L97) updates $\boldsymbol{R}_1$ and $\boldsymbol{R}_2$ (and returns them in a concatenated vector). The posterior mean and covariance are calculated as a function of those quantities in [**select_poisson**](https://github.com/cran/pogit/blob/b0f4b72eaabb1ce32efa1f4ce5517be7ece2afd0/R/select_poisson.R#L110), where also a sampler from the posterior is drawn. The two additional functions [**mixcomp_poisson**](https://github.com/cran/pogit/blob/master/R/mixcomp_poisson.R) and [**compute_mixture**](https://github.com/cran/pogit/blob/b0f4b72eaabb1ce32efa1f4ce5517be7ece2afd0/R/compute_mixture.R) take care of the finite mixture approximation. **mixcomp_poisson** holds the necessary parameters for the finite mixture representation in the Poisson case, and **compute_mixture** uses those to calculate the approximation. The **mixcomp_poisson** function is (essentially) just a list, so I will ignore that for now and just call it from the package when I need it.

All these functions are present in JAGS as well, in: [AuxMixPoisson.CC](https://github.com/todesking/JAGS-code/blob/master/src/modules/glm/samplers/AuxMixPoisson.cc) and [AuxMixPoisson.h](https://github.com/todesking/JAGS-code/blob/master/src/modules/glm/samplers/AuxMixPoisson.h), and [LGMix.CC](https://github.com/todesking/JAGS-code/blob/master/src/modules/glm/samplers/LGMix.cc).

### Updating the inter-arrival times $\boldsymbol{\tau}$

Code from the **pogit** package includes R-functions that are challenging to incorporate in NIMBLE, so I built my functions from the ground up based on their code.

Here is a functiont to update $\boldsymbol{\tau}$:

```{r tau_func}
library(nimble)
update_tau = function(response.nodes = integer(1), n.response.nodes = integer(), n.zero.response.nodes = integer(), lambda = double(1)){
                        taunew = matrix(0, nrow = n.response.nodes, ncol = 2)
                        taunew[,1] <- rexp(n.response.nodes, lambda)
                        tau2 <- rbeta(n.response.nodes - n.zero.response.nodes, response.nodes[response.nodes>0], 1)
                        taunew[response.nodes > 0,  1] = 1-tau2 + taunew[response.nodes > 0,1]
                        taunew[response.nodes > 0,  2]  = tau2
                        taunew[response.nodes == 0, 1] = 1 +  taunew[response.nodes == 0, 1]
                        return(taunew)
                      }
```

I can simulate some data following a simple example:
```{r sim_dat}
N <- 1000 # number of observations
beta0 <- 1  # intercept
beta1 <- 1  # slope
set.seed(1);x <- rnorm(n=N)  # standard normal predictor
eta <- beta0*1 + beta1*x  # linear predictor function
lambda <- exp(eta)  # link function
set.seed(1);y <- rpois(n=N, lambda=lambda)  # Poisson DV
```

and verify that it returns the same result as **pogit**:
```{r check_tau}
# calculate tau
set.seed(1);my_tau <- update_tau(response.nodes = y, n.response.nodes = N, n.zero.response.nodes = sum(y==0), lambda = lambda) # returns a matrix
# pogit needs these objects already here, they include properties of the data
mcomp <- pogit:::mixcomp_poisson()
compmix.pois <- pogit:::get_mixcomp(y = y, mcomp = mcomp)
set.seed(1);pogit_tau <- pogit:::iams1_poisson(y = y, mu = lambda, compmix.pois = compmix.pois) # returns a list of t1, t2

# calculate differences between vectors to verify correctness
sum(my_tau[,1] - pogit_tau$t1) # should be zero
sum(my_tau[y>0,2] - pogit_tau$t2) # should be zero
```

### Updating the indicator variables $\boldsymbol{R}_1$ and $\boldsymbol{R}_2$

So, tau is correct. Let us continue with a function for the indicator variables. The **pogit** package returns the two variables from the **iams2_poisson** function in a concatenated string, I have written two separate functions:

```{r R1_func}
update_R1 = function(n.response.nodes = integer(), lp = double(1), t1 = double(1), mcompm = double(2), mcompv = double(2), c1 = double(2))
                      {
                        minlogt1minlp <- matrix(-log(t1) - lp,ncol=10,nrow=n.response.nodes)
                        repmcompm  <- t(matrix(mcompm[1,], ncol = n.response.nodes, nrow = 10))
                        repmcompv <- t(matrix(mcompv[1,], ncol = n.response.nodes, nrow = 10))
                        rgm   <- c1 - 0.5*(minlogt1minlp - repmcompm)^2/repmcompv
                        for(i in 1:10){ #ncol rgm
                          rgm[rgm[,i]==0,i] <- -Inf
                        }
                        mx <- numeric(n.response.nodes)
                        e1 <- matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:n.response.nodes){
                          mx[i] <- max(rgm[i,])
                          if(mx[i]==-Inf)mx[i]=0
                          e1[i,] <- exp(rgm[i,]-mx[i])
                        }
                        rgmod <- numeric(n.response.nodes)
                        for(i in 1:n.response.nodes){
                          rgmod[i] <- sum(e1[i,])
                        }
                        e1.new = matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:10){
                          e1.new[,i] <- e1[,i]/rgmod # for safety might want to check for 0/0 i.e., nan
                        }
                        tri.mat = matrix(1, 10, 10)
                        for (j in 1:9) {
                          for (i in (j+1):10) {
                            tri.mat[i,j] = 0
                          }
                        }
                        Fn    <- e1.new%*%tri.mat
                        
                        # determination of random indicators R1
                        u <- runif(n.response.nodes, 0, 1)
                        R1.temp <- matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:10){
                          R1.temp[,i] <- u<Fn[,i] #can probably replace this step by "sample()"?
                        }
                       R1 <- numeric(n.response.nodes)
                        for(i in 1:n.response.nodes){
                          R1[i] <- 11-sum(R1.temp[i,])
                        }
                       return(R1)
                      }
```
```{r R2_func}
update_R2 = function(response.nodes = integer(1), n.response.nodes = integer(), n.zero.response.nodes = integer(), lp = double(1), t2 = double(1), vy = double(2), my = double(2), wy = double(2)){
                        vy2 <- vy
                        for(i in 1:10){
                          vy2[vy2[,i]==0,i]<-1
                        }
                        lwy <- wy
                        for(i in 1:10){
                          lwy[lwy[,i]==0,i] <- 1
                          lwy[lwy[,i]<0,i] <- 1
                          lwy[,i] <- log(lwy[,i])
                        }
                        lvy <- vy
                        for(i in 1:10){
                          lvy[lvy[,i]==0,i] <- 1
                          lvy[lvy[,i]<0,i] <- 1
                          lvy[,i] <- log(lvy[,i])
                        }
                        c2 <- (lwy - 0.5*lvy) 
                        minlogt2minlp <- matrix(-log(t2) - lp[response.nodes>0],ncol=10,nrow=n.response.nodes-n.zero.response.nodes)
                        
                        kill <- matrix(vy > 0, n.response.nodes - n.zero.response.nodes, 10)
                        xx     <- minlogt2minlp*kill
                        rgmx   <- c2 - 0.5*(xx - my)^2/vy2
                        for(i in 1:10){ #ncol r gm
                          rgmx[rgmx[,i]==0,i] <- -Inf
                        }
                        e2 <- matrix(0,nrow = n.response.nodes-n.zero.response.nodes, ncol = 10)
                        mx2 = numeric(n.response.nodes-n.zero.response.nodes)
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          mx2[i] = max(rgmx[i,])
                          if(mx2[i]==-Inf)mx2[i]=0
                          e2[i,] <- exp(rgmx[i,]-mx2[i])
                        }
                        rgmodx <- numeric(n.response.nodes-n.zero.response.nodes)
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          rgmodx[i] <- sum(e2[i,])
                        }
                        e2.new = matrix(0,nrow=n.response.nodes-n.zero.response.nodes, ncol = 10)
                        for(i in 1:10){
                          e2.new[,i] <- e2[,i]/rgmodx # for safety might want to check for 0/0 i.e., nan
                        }
                        
                        tri.mat = matrix(1, 10, 10)
                        for (j in 1:9) {
                          for (i in (j+1):10) {
                            tri.mat[i,j] = 0
                          }
                        }
                        
                        Fx     <- e2.new%*%tri.mat
                        
                        # determination of random indicators R2
                        ux <- runif(n.response.nodes - n.zero.response.nodes, 0, 1)
                        R2.temp <- matrix(0, ncol = 10, nrow = n.response.nodes - n.zero.response.nodes)
                        for(i in 1:10){
                          R2.temp[,i] <- ux<Fx[,i]
                        }
                        
                        R2 <- numeric(n.response.nodes-n.zero.response.nodes)
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          R2[i] <- 11-sum(R2.temp[i,])
                        }
                        return(R2)
                      }
```

Comparing again to the **pogit** package:

```{r check_R}
# first I compute a constant that both functions need
c1 <- t(matrix(log(mcomp$w[1, ]) - 0.5*log(mcomp$v[1, ]), nrow = 10, ncol = N))
# calculate my Rs
set.seed(1);my_R1 <- update_R1(n.response.nodes = N, lp = eta, t1 = pogit_tau$t1, mcompm = mcomp$m, mcompv = mcomp$v, c1 = c1)
set.seed(1);my_R2 <- update_R2(response.nodes = y, n.response.nodes = N, n.zero.response.nodes = sum(y==0), lp = eta, t2 = pogit_tau$t2, vy = compmix.pois$vy, my = compmix.pois$my, wy = compmix.pois$wy)
# pogit Rs
cm1 <- list(comp = list(m = mcomp$m[1, ], v = mcomp$v[1, ], w = mcomp$w[1, ]), c1 = c1) # pogit requires a list of arguments
set.seed(1);pogit_R <- pogit:::iams2_poisson(n = N, tau1 = pogit_tau$t1, tau2 = pogit_tau$t2, logMu = eta, logMugz = eta[y>0], cm1 = cm1, compmix = compmix.pois)

sum(pogit_R[1:N]-my_R1) # should be zero
sum(pogit_R[-c(1:N)]-my_R2) # should be zero 
```

the output for both sums should be zero, but for $\boldsymbol{R}_2$ is not. I suspect this is because the seed for *pogit_R* is not set correctøy, since the function includes two random number generators and mine only one. However, I have verified the calculation for $\boldsymbol{R}_2$ and it is correct.

### Posterior moments

All that remains is to calculate the posterior moments and generate a sample from the posterior. Unfortunately, this is harder to reproduce with the **pogit** package because there is no function that returns those quantities. In an attempt to verify my output anyway, I have written a function that outputs the posterior mean, covariance, and a proposal, based on the code in **select_poisson.R**:

```{r pogit_post_calc}
pogit_post_calc <- function(X, compmix, tau1, tau2, R, cm1){
X <- as.matrix(X)
n <- nrow(X)
# mixture component means and variances
m1 <- cm1$comp$m[R[1:n]]
m2 <- compmix.pois$my[cbind(seq_len(compmix.pois$ngz), R[-(1:n)])]
mR <- as.matrix(c(m1,m2), (n + compmix.pois$ngz))
v1 <- cm1$comp$v[R[1:n]]
v2 <- compmix.pois$vy[cbind(seq_len(compmix.pois$ngz), R[-(1:n)])]
invSig <- 1/sqrt(c(v1,v2))

# stacking and standardizing
tauS <- c(tau1, tau2)
xS <- rbind(X, X[compmix.pois$igz, , drop = FALSE])
yS <- (-log(tauS) - mR)*invSig

Xall <- xS*kronecker(matrix(1, 1, ncol(X)), invSig) # X times inv. sd.dev of mixtures

# inv. prior variance
invA0 <- diag(ncol(Xall))
a0 = rep(0,ncol(Xall)) # prior mean

AP    <- solve(invA0 + t(Xall)%*%Xall)  # A = (A0^-1 + (Z*)'Sigma^-1 Z*)
aP    <- AP%*%(invA0%*%a0 + t(Xall)%*%yS) # a = A(A0^-1*a0 + (Z*)'Sigma^-1*y
zetaP <- t(chol(AP))%*%matrix(rnorm(ncol(Xall)), ncol(Xall), 1) + aP 
return(list(postMean = aP, postCov = AP, proposal = zetaP))
}
```

and my function, only for testing purposes here:

```{r my_post_calc}
my_post_calc <- function(X, n.response.nodes, n.zero.response.nodes, response.nodes, R1, R2, tau, mcompv, mcompm, my, vy){
    X <- as.matrix(X)
    n.param.nodes <- ncol(X)
    m1 <- mcompm[1,R1]
    m2 <- numeric(n.response.nodes-n.zero.response.nodes)
    for(i in 1:(n.response.nodes-n.zero.response.nodes)){
      m2[i] <- my[i,R2[i]]
    }
    # second: mixture component variances
    v1 <- mcompv[1,R1]
    v2 <- numeric(n.response.nodes-n.zero.response.nodes)
    for(i in 1:(n.response.nodes-n.zero.response.nodes)){
      v2[i] <- vy[i,R2[i]]
    }
    invSigS = numeric(n.response.nodes)
    ys1 = (-log(tau[,1])-m1)/sqrt(v1) # this is also "AuxMixPoisson:value()" in JAGS
    ys2 = (-log(tau[response.nodes>0,2])-m2)/sqrt(v2)
    invSigS = 1/v1
    invSigS[response.nodes>0] = invSigS[response.nodes>0]+1/v2
    # prior parameters
    bPriorCov = diag(n.param.nodes)
    bPriorMean = rep(0, n.param.nodes)
    # posterior parameters
    bPostCov = solve(solve(bPriorCov, diag(n.param.nodes)) + t(X)%*%diag(invSigS)%*%X ,diag(2))
    bPostMean = bPostCov%*%(solve(bPriorCov,diag(n.param.nodes))%*%bPriorMean+ (t(X)%*%diag(1/sqrt(v1))%*%ys1+t(X[response.nodes>0,])%*%diag(1/sqrt(v2))%*%ys2)) # see eq (8) Fruhwirth-Schnatter and Wagner 2006 or L109-111 from pogit::select_poisson.Rs
    # a sample from the posterior
    proposal <- rmnorm_chol(1, mean = c(bPostMean), cholesky = chol(bPostCov), prec_param = 0)
return(list(postMean = bPostMean, postCov = bPostCov, proposal = proposal))
}
```

test it:
```{r test_post_calc}
set.seed(1);my_post <- my_post_calc(cbind(1,x), N, sum(y==0), y, pogit_R[1:N], pogit_R[-c(1:N)], my_tau, mcomp$v, mcomp$m, compmix.pois$my, compmix.pois$vy)
set.seed(1);pogit_post <- pogit_post_calc(cbind(1,x),compmix,pogit_tau$t1, pogit_tau$t2, pogit_R, cm1)
sum(my_post$postMean-pogit_post$postMean)
sum(my_post$postCov-pogit_post$postCov)
```

### Putting it all together in NIMBLE

Finally, we put everything together in a NIMBLE sampler;

```{r nim_samp}
# sampler for Poisson responses based on Fruhwirth-Schnatter and Wagner 2006 and Fruhwirth-Schnatter et al. 2009
# as implemented in the JAGS and the pogit packages
sampler_glm_pois <- nimbleFunction(
  
  contains = sampler_BASE,
  name = 'sampler_glm_pois',
  setup = function(model, mvSaved, target, control) {
    # get response data nodes
    response.nodes.names <- names(which(model$isStoch(names(model$origData)))) # names of response nodes
    if(!all(model$getDistribution(response.nodes.names)=="dpois"))stop("This sampler is designed only for Poisson responses.")
    response.name  = unique(gsub("\\[.*","",response.nodes.names))
    response.nodes <- values(model, response.nodes.names) # response data
    n.response.nodes <- length(response.nodes) # number of observations
    n.zero.response.nodes = sum(response.nodes==0) # number of zero observations
    ## get distribution parameter names
    dist.par.names = model$getParents(response.nodes.names,determOnly = F, immediateOnly = T)
    distpar.name  = unique(gsub("\\[.*","",dist.par.names))
    lp.nodes.names = model$getParents(dist.par.names,immediateOnly = T, determOnly = T,includeData = F)
    lp.name  = unique(gsub("\\[.*","",lp.nodes.names))
    # get corresponding "data" nodes to target, should do this automatically based o "target"
    X <- model$X
    # number of target parameters
    n.param.nodes <- dim(X)[2]
    # prior parameters, should get these based on "target"
    bPriorMean = numeric(n.param.nodes) # still replace if different
    bPriorCov = diag(n.param.nodes) # still replace if different
    # mixture components
    # v: variances (s), m: means (m), w: weights
    mcomp <- if(!is.null(control$mcomp)) control$mcomp else mcomp <- pogit:::mixcomp_poisson()
    mcompm <- mcomp$m
    mcompv <- mcomp$v
    mcompw <- mcomp$w
    # initialize R, indicator variables
    R1 <- rep(NA_real_, n.response.nodes)
    R2 <- rep(NA_real_, n.response.nodes-n.zero.response.nodes)
    tau <- matrix(NA_real_, n.response.nodes,ncol=2)
    # start by getting mixture components, this is "get_mixcomp" in pogit
    # m: means for mixture components, v: scales, w: weights
    if(sum(response.nodes<3e4)==n.response.nodes){
      my = mcomp$m[response.nodes[response.nodes>0],]
      vy = mcomp$v[response.nodes[response.nodes>0],]
      wy = mcomp$w[response.nodes[response.nodes>0],]
    }else{
      wy = vy = my = matrix(0, nrow = n.response.nodes - n.zero.response.nodes, ncol = 10)
      my[response.nodes[response.nodes>0] <= 3e4,1:dim(mcomp$m)[2]] = mcomp$m[response.nodes[(response.nodes>0) & (response.nodes <= 3e4)],]
      vy[response.nodes[response.nodes>0] <= 3e4,1:dim(mcomp$v)[2]] = mcomp$v[response.nodes[(response.nodes>0) & (response.nodes <= 3e4)],]
      wy[response.nodes[response.nodes>0] <= 3e4,1:dim(mcomp$w)[2]] = mcomp$w[response.nodes[(response.nodes>0) & (response.nodes <= 3e4)],]
      # this comes from "compute_mixture.R"
      for(i in response.nodes[response.nodes>3e4]){
        my[response.nodes[response.nodes>0]==i,1] <- -digamma(i)
        vy[response.nodes[response.nodes>0]==i,1] <- trigamma(i)
        wy[response.nodes[response.nodes>0]==i,1] <- 1 
      }
    }
    vy[is.na(vy)] <- 0
    my[is.na(my)] <- 0
    wy[is.na(wy)] <- 0
    # for use in step 3
    c1 <- t(matrix(log(mcompw[1, ]) - 0.5*log(mcompv[1, ]), nrow = 10, ncol = n.response.nodes)) # not sure what this is yet
    
    calcNodes <- model$getDependencies(target)
    model$simulate()
    model$calculate(calcNodes)
    targetAsScalar <- model$expandNodeNames(target, returnScalarComponents = TRUE)
  },
  
  run = function() {
    # 1. simulate parameters given R, tau
    # this part is covered in "select_poisson.R"
    if(!any_na(R1)){ # I don't update parameters on the first iteration as to generate sensible "warm" start for R1,R2,t1,t2
    # first mixture component means
    m1 <- mcompm[1,R1]
    m2 <- numeric(n.response.nodes-n.zero.response.nodes)
    for(i in 1:(n.response.nodes-n.zero.response.nodes)){
      m2[i] <- my[i,R2[i]]
    }
    # second: mixture component variances
    v1 <- mcompv[1,R1]
    v2 <- numeric(n.response.nodes-n.zero.response.nodes)
    for(i in 1:(n.response.nodes-n.zero.response.nodes)){
      v2[i] <- vy[i,R2[i]]
    }
    invSigS = numeric(n.response.nodes)
    ys1 = (-log(tau[,1])-m1)/sqrt(v1) # this is also "AuxMixPoisson:value()" in JAGS
    ys2 = (-log(tau[response.nodes>0,2])-m2)/sqrt(v2)
    invSigS = 1/v1
    invSigS[response.nodes>0] = invSigS[response.nodes>0]+1/v2
    v1mat = matrix()
    # posterior parameters
    bPostCov = solve(solve(bPriorCov, diag(n.param.nodes)) + t(X)%*%diag(invSigS)%*%X ,diag(2))
    bPostMean = bPostCov%*%(solve(bPriorCov,diag(n.param.nodes))%*%bPriorMean+ t(X)%*%diag(1/sqrt(v1))%*%ys1+t(X[response.nodes>0,])%*%diag(1/sqrt(v2))%*%ys2) # see eq (8) Fruhwirth-Schnatter and Wagner 2006 or L109-111 from pogit::select_poisson.Rs
    # a sample from the posterior
    proposal <- rmnorm_chol(1, mean = c(bPostMean), cholesky = chol(bPostCov), prec_param = 0)
    values(model, targetAsScalar) <<-  proposal
    # update lp, lambda after calculating proposal
    model$calculate(calcNodes)
    }
    ## get LP and mu
    lp = values(model, lp.nodes.names)
    lambda = values(model, dist.par.names)
    #2. update tau's, step 1 in Fruhwirth-Schnatter et al. 2009
    # this part is covered in dataug_pois_iams.R->iams1_poisson
    # tau are latent variables that represent the inter-arrival times of a Poisson process
    # get_mixcomp_poisson in dataug_pois_iams.R just gets components that we need for the other functions
    
    update_tau(response.nodes, n.response.nodes, n.zero.response.nodes, lambda)
    t1 <- tau[,1]
    t2 <- tau[response.nodes>0,2]
    # 3. update R's, step 2 in Fruhwirth-Schnatter et al. 2009
    # this part is covered in dataug_pois_iams.R->iams2_poisson
    # Rmix are normal mixture approximations to Poissons
    
    update_R1(n.response.nodes, lp, t1, mcompm, mcompv, c1)
    update_R2(response.nodes, n.response.nodes, n.zero.response.nodes, lp, t2, vy, my, wy)
    # keep the model and mvSaved objects consistent
    copy(from = model, to = mvSaved, row = 1, 
         nodes = target, logProb = TRUE)
  },
  
  methods = list(update_tau = function(response.nodes = integer(1), n.response.nodes = integer(), n.zero.response.nodes = integer(), lambda = double(1)){
                        taunew = matrix(0, nrow = n.response.nodes, ncol = 2)
                        taunew[,1] <- rexp(n.response.nodes, lambda)
                        tau2 <- rbeta(n.response.nodes - n.zero.response.nodes, response.nodes[response.nodes>0], 1)
                        taunew[response.nodes > 0,  1] = 1-tau2 + taunew[response.nodes > 0,1]
                        taunew[response.nodes > 0,  2]  = tau2
                        taunew[response.nodes == 0, 1] = 1 +  taunew[response.nodes == 0, 1]
                        tau<<-taunew
                      },
                    update_R1 = function(n.response.nodes = integer(), lp = double(1), t1 = double(1), mcompm = double(2), mcompv = double(2), c1 = double(2))
                      {
                        minlogt1minlp <- matrix(-log(t1) - lp,ncol=10,nrow=n.response.nodes)
                        repmcompm  <- t(matrix(mcompm[1,], ncol = n.response.nodes, nrow = 10))
                        repmcompv <- t(matrix(mcompv[1,], ncol = n.response.nodes, nrow = 10))
                        rgm   <- c1 - 0.5*(minlogt1minlp - repmcompm)^2/repmcompv
                        for(i in 1:10){ #ncol rgm
                          rgm[rgm[,i]==0,i] <- -Inf
                        }
                        mx <- numeric(n.response.nodes)
                        e1 <- matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:n.response.nodes){
                          mx[i] <- max(rgm[i,])
                          if(mx[i]==-Inf)mx[i]=0
                          e1[i,] <- exp(rgm[i,]-mx[i])
                        }
                        rgmod <- numeric(n.response.nodes)
                        for(i in 1:n.response.nodes){
                          rgmod[i] <- sum(e1[i,])
                        }
                        e1.new = matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:10){
                          e1.new[,i] <- e1[,i]/rgmod # for safety might want to check for 0/0 i.e., nan
                        }
                        tri.mat = matrix(1, 10, 10)
                        for (j in 1:9) {
                          for (i in (j+1):10) {
                            tri.mat[i,j] = 0
                          }
                        }
                        Fn    <- e1.new%*%tri.mat
                        
                        # determination of random indicators R1
                        u <- runif(n.response.nodes, 0, 1)
                        R1.temp <- matrix(0, ncol = 10, nrow = n.response.nodes)
                        for(i in 1:10){
                          R1.temp[,i] <- u<Fn[,i] #can probably replace this step by "sample()"?
                        }
                        
                        for(i in 1:n.response.nodes){
                          R1[i] <<- 11-sum(R1.temp[i,])
                        }
                      },
                    update_R2 = function(response.nodes = integer(1), n.response.nodes = integer(), n.zero.response.nodes = integer(), lp = double(1), t2 = double(1), vy = double(2), my = double(2), wy = double(2)){
                      # this function only works if l380-384 are here instead of in "setup" for some reason
                      # otherwise it returns NaN..
                        vy2 <- vy
                        for(i in 1:10){
                          vy2[vy2[,i]==0,i]<-1
                        }
                        lwy <- wy
                        for(i in 1:10){
                          lwy[lwy[,i]==0,i] <- 1
                          lwy[lwy[,i]<0,i] <- 1
                          lwy[,i] <- log(lwy[,i])
                        }
                        lvy <- vy
                        for(i in 1:10){
                          lvy[lvy[,i]==0,i] <- 1
                          lvy[lvy[,i]<0,i] <- 1
                          lvy[,i] <- log(lvy[,i])
                        }
                        c2 <- (lwy - 0.5*lvy)
                        
                        minlogt2minlp <- matrix(-log(t2) - lp[response.nodes>0],ncol=10,nrow=n.response.nodes-n.zero.response.nodes)
                        
                        kill <- matrix(vy > 0, n.response.nodes - n.zero.response.nodes, 10)
                        xx     <- minlogt2minlp*kill
                        
                        rgmx   <- c2 - 0.5*(xx - my)^2/vy2
                        for(i in 1:10){ #ncol r gm
                          rgmx[rgmx[,i]==0,i] <- -Inf
                        }
                        e2 <- matrix(0,nrow = n.response.nodes-n.zero.response.nodes, ncol = 10)
                        mx2 = numeric(n.response.nodes-n.zero.response.nodes)
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          mx2[i] = max(rgmx[i,])
                          if(mx2[i]==-Inf)mx2[i]=0
                          e2[i,] <- exp(rgmx[i,]-mx2[i])
                        }
                        rgmodx <- numeric(n.response.nodes-n.zero.response.nodes)
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          rgmodx[i] <- sum(e2[i,])
                        }
                        e2.new = matrix(0,nrow=n.response.nodes-n.zero.response.nodes, ncol = 10)
                        for(i in 1:10){
                          e2.new[,i] <- e2[,i]/rgmodx # for safety might want to check for 0/0 i.e., nan
                        }
                        
                        tri.mat = matrix(1, 10, 10)
                        for (j in 1:9) {
                          for (i in (j+1):10) {
                            tri.mat[i,j] = 0
                          }
                        }
                        
                        Fx     <- e2.new%*%tri.mat
                        
                        # determination of random indicators R2
                        ux <- runif(n.response.nodes - n.zero.response.nodes, 0, 1)
                        R2.temp <- matrix(0, ncol = 10, nrow = n.response.nodes - n.zero.response.nodes)
                        for(i in 1:10){
                          R2.temp[,i] <- ux<Fx[,i]
                        }
                        
                        for(i in 1:(n.response.nodes-n.zero.response.nodes)){
                          R2[i] <<- 11-sum(R2.temp[i,])
                        }
                      },
                 reset = function () {}
                    )
)
```

write some BUGS code for a simple Poisson regression;

```{r nim_script}
library(nimble)
pois.glm <- nimbleCode({
  ## Likelihood
  for(i in 1:N){
    y[i] ~ dpois(lambda[i])
    log(lambda[i]) <- eta[i]
    eta[i] <- inprod(beta[1:k],X[i,])
  }     
  ## Priors 
  for(i in 1:k){
    beta[i] ~ dnorm(0,1)
  }
})
```

set-up the model:

```{r setup_model}
dat <- list(X=cbind(1,x),  # predictors
                y=y)  # DV
const <- list(
                N=N,  # sample size
                k =ncol(dat$X)) # number of covariates,
# construct model object
model <- nimbleModel(pois.glm, const, dat, inits = list(beta = c(1,1)))
mod <- nimble::compileNimble(model)
```

and add and run it for a few iterations to see that we get something sensible:

```{r run_mcmc}
nimbleMCMCconf <- configureMCMC(mod, monitors = c("beta"), print = FALSE)
nimbleMCMCconf$removeSampler("beta")
nimbleMCMCconf$addSampler("beta",type="sampler_glm_pois", print = T)
nimbleMCMCb <- buildMCMC(nimbleMCMCconf)
nimbleMCMCc <- compileNimble(nimbleMCMCb, project = mod)
test_run <- runMCMC(nimbleMCMCc, niter = 1000, nburnin = 500, thin = 1, nchains = 3, samplesAsCodaMCMC = TRUE)
summary(test_run)# wrong answer :(
```
